The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-05-27 Paper/jaccardbaetica.tex
comentarios
new changes xavi correction

The aim of this study is to explore economics dynamics in the production and distribution of olive oil trade. 
Our case of study has been focused on the production processes located in \textit{Baetica} province (currently Andalusia) from 1st to 3rd AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. \textit{Baetica} became an important production and distribution centre during the Roman Empire. However, it remains under debate about how this province was organised and whether it could be possible to identify patterns in the olive oil market. Amphoric stamps are used to identify the presence of different groups that might share similar stamps. To achieve this goal, we analyse a set of stamps from two centres: 1) producer centres by analysing different workshops in \textit{Baetica} province and 2) two Roman provinces such as \textit{Germania} and \textit{Britannia} as a receptor centres. They will be used to detect a connection between the distribution of amphoric stamps and the economic structure in both centres. Here, we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance.

@@

The role of the Mediterranean sea for the trade had an enormous impact during the Roman Empire. The spread of thousands of different goods along maritime shipping allowed a progressive exploitation of communication networks and a major interaction between communities \citep{rodriguez_baetican_1998, 
temin_market_2001,
bevan_mediterranean_2014}. The exchange of different products between regions can be seen in both archaeological and written records as a frequent indicator of Roman trade.

@@

Given this fact, an important mechanism of production control under the Roman government was created spreading by different areas attending to the richness of different places. Roman provinces such as \textit{Baetica} (currently Andalusia, south Spain) became important production centres of olive oil during the Roman Empire. Olive oil was considered the liquid gold since it was used in different aspect of daily life for cooking, hygiene or lighting  \citep{mattingly_d.j._oil_1988}. As a consequence, this product was distributed in a large number of amphorae across the provinces, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

The understanding of the olive oil production processes has been widely discussed over many decades\citep{rodriguez_economioleicola_1977, Chic_hispania_1997,millet_anforas_1998}. Despite this topic remains under debate: what patterns were followed for the distribution of olive oil to the different provinces?
Did each province follow a different pattern for its distribution? Nor the lack of written records has made possible to detect any indication of patterns in the olive oil market.

Advances in the research of the Roman world have currently led an environment with more diverse commercial dynamics \citep{duncan1982economy,
temin_economy_2006,
quantifyingwilson2009}. The application of different quantitative approaches combined with the archaeological evidence has allowed us partly to deal with the complexity of the Roman production \citep{brughmans_roman_2016,
orengo_seeds_2016,bayesian_2018,
coto-sarmiento_identifying_2018,
rubio-campillo_ecology_2018}.

%\xavi{le das demasiada importancia al similarity index. Es una simple herramienta de tu trabajo, y no un objetivo}.
This paper aims to study the olive oil market connection between provinces by calculating the similarity of stamps. Specifically, our work pretends to detect microeconomic processes focused on a commercial product from a specific province \citep{isaksen_network_2006}. We want to understand the pattern of olive oil production linked to amphora workshops and amphoric stamps used to mark them. We focus here on exploring the economic relation between stamps and amphora production and distribution centres.

Two case studies have been studied in order to analyse the relation between production centres (\textit{Baetica}) and consumption centres (\textit{Britannia} and \textit{Germania}). In the case of \textit{Baetica} province, we want to identify the role of the stamps in the organisation of the workshop; in Roman provinces, our aim is detecting groups of stamps concentrated in an area or if some groups have an important role for the exportation of olive oil in those provinces. This economic connection could be identified by different aspects: a) correlation between spatial distance and centres based on the idea that closer workshops concentrate similar amphoric stamps in a specific area than the distant workshops and b) groups of similar stamps were concentrated in a specific Roman province.

In particular, we study the distribution of amphoric stamp in order to identify a correlation between geographical distance and amphoric stamps. Based on this assumption, we proposed three hypotheses: a) we can identify a correlation between spatial distance and the distribution of stamps, b) stamps located in close workshops share similar traits and c) Low mobility of amphoric stamps to other regions: stamps always stay in the same region.

@@

The high demand of Roman provinces stimulated by the good condition of the \textit{Baetica} lands allowed to develop a massive infrastructure of olive oil production. The production and distribution of olive oil were growing exponentially for almost three centuries \citep{remesal_concierto}. As a result, hundreds of amphora workshops were made to support this high demand of the Roman Empire. The amphora workshops were situated in different locations in \textit{Baetica} province, along the rivers Guadalquivir and Genil (see Fig.\ref{workshop}).

The location of the workshops with a riverine connectivity allowed to arrange a connection with the Mediterranean and Atlantic routes \citep{garcia_vargas_enrique_formal_2010}.

@@

\section{A potential indicator of Roman economy: Dressel 20 stamps}
%\xavi{no sería mejor que esta sección esté antes de hablar de los casos de estudios? Así, podrías ya hablar de sellos encontrados en germania y britannia}

@@

Most stamps display a large activity of production that it can be difficult for specifying an accurate chronology in some cases. They have been frequently dated with the consular dating by studying \textit{tituli picti} found in \textit{Monte Testaccio} \citep{Testaccio1, berni_millet_epigrafianforica_2008}.
However, the chronology of stamps could be also biased in other sites taking into account that amphorae were deposited after being produced and consumed.  
\maria{no sé cómo explicarlo mejor ni como citarlo porque no hay citas}

@@

Frequently, they were mainly marked in handles, but seldom in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion. Stamps are mostly formed by a code of three letters and they can appear in abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}. Despite researchers identify stamps as identity marks, there is not a general consensus about the meaning of the stamps \citep{rodriguez_baetican_1998}. Alike most Dressel 20 containers were not stamped being difficult to determinate.

@@

Nevertheless, some challenges remain under discussion such as how this production was organised and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow a distribution pattern? Did stamps share the same workshop? Neither written records have been found that it could explain the economic role of \textit{Baetica} province in the Roman organisation.

@@

The conquest of new provinces allowed the Roman Empire the arrival of new resources through Mediterranean and Atlantic routes. This lead to a progressive change in the economic and social structure creating new trade networks for supplying Roman conquest. As a result, Augusto's administration created the figure of the \textit{praefectura annonae} for the supply of wheat. The role of the \textit{praefectura annonae} was mainly to provide though \textit{frumentationes} a quantity per month of wheat to Roman citizens \citep{remesal_annona_1986,remesal_concierto}.

@@

The military consumption of olive oil in Roman provinces could be related to two aspects: a) cultural consumption whose product is consumed by cultural reasons such as to identity or habit and b) economical reasons where olive oil distribution is based on the transport costs, unlike other products \citep[69-70]{carreras_britannia_1998}.

@@

At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}. This fact will be spatial incidence in sites close at Hadrian Wall.  
Olive oil production in \textit{Baetica} would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic points \citep{carreras_atlantic_2012}. The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. Thus, the network was mainly focused on the support of soldiers during the military campaigns.

@@

Finally, The increase in olive oil exports would experience a progressive slowdown from the third century A.D., coinciding with the change in market strategy in the Empire. As a result, a progressive decrease of Dressel 20 is documented being gradually replaced for Dressel 23 amphorae \citep{rodriguez1991aceite,millet_anforas_1998}.

@@

As the case of \textit{Britannia}, the consumption of olive oil in \textit{Germania} was not detected before the conquest. This is evidenced by the presence of Dressel 20 amphora in different sites close to limes.

However, the Roman studies about the presence of Dressel 20 amphorae in \textit{Germania} has not had the same repercussion as the rest of the province. This could be explained due to the lack of archaeological sources \citep{horacio2010llegada}.

@@

The presence of the Roman army encouraged the exchange in the province showed by the arrival of this product both civil settlement and military sites with a mayor concentration at German limes. It seems that some Baetican centres would be assigned to the support of olive oil. However, this hypothesis can not be proven by the lack of archaeological sources \citep[125]{remesal_concierto}.

\section{Material}

%\xavi{estas 2 cosas nunca van juntas. En Method explicas las técnicas, y Analysis en general no se usa porque es el todo del trabajo}

The goal of this study is exploring the effect of the production patterns between different centres. We are specially interested in identifying links between producers and consumption centres by using amphoric stamps. To do that, we use the CEIPAC database to collect the stamps from different places. The CEIPAC dataset contains over 50.000 of epigraphy records found in amphorae, mostly from \textit{Monte Testaccio}. Therefore, this study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. A way to analyse is to use a quantitative framework to measure the similarity between amphoric stamps. Here we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potentially spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance.

@@

We studied a dataset of 3798 stamps collected from different Dressel 20 amphora workshops in \textit{Baetica} province. The stamp database was compiled by CEIPAC group \citep{remesal_centro_2015} (see CEIPAC database here \url{http://romanopendata.eu}). However, approximately the 70 \% of stamps cannot be tested due to fragmentation or incomplete information. Consequently, we discard integrate the fragmented stamps in our dataset. We finally filtered a total sample of 987 stamps composed of 130 different stamps from 81 workshops.

@@

From the database, we collected the site where stamps were found and the stamp code. We also created a new column with the area of the stamps. These areas, known as \textit{conventus}, were administrative centres for territorial organization in the Roman Empire. Dressel 20 stamps were found in three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter Hispalis), \textit{Cordubensis} (currently C\'ordoba, hereafter Corduba) and \textit{Astigi} (currently Écija, Sevilla, hereafter Astigi) \citep{rodriguez_economioleicola_1977,
chicdatos2001,berni_millet_epigrafianforica_2008} .

We calculated the frequency of distribution in amphora stamps with the purpose to carry out an exploratory analysis (EDA) to know the distribution and the number of stamps per each centre.

The frequency of stamps per workshop can be seen in Fig. \ref{stamps}. Most 
workshop contained only one stamp whereas one workshop (La Catria) concentrated a high frequency of amphoric stamps with a total number of 228 stamps. The type of distribution is also frequent in amphora production where we observed a self-organized complex system pattern with a major concentration of the number of stamps in a few workshops \citep{bayesian_2018,coto-sarmiento_identifying_2018}.

@@

The distribution of amphora stamps in different \textit{conventus} can be seen in Fig. \ref{frequency}. The majority of stamps found are concentrated in \textit{Hispalis} with 574 stamps while \textit{Corduba} and \textit{Astigi} with 267 and 146 stamps, respectively. Mostly, workshops show a homogeneity on the frequency of stamps except for both La Catria and Arva that exhibit a big amount of stamps with 29 different code stamps. According to previous studies, those workshops became in most important centres of amphora production although they could have more intensely prospected than others. \citep{arva_1997}.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/frequency}
\caption{Distribution of the number of different code stamps (X axis) for each \textit{conventus} (Y axis). Each dot corresponds to a workshop sorted by different areas. Colors are represented by areas divided into Hispalis (red), Astigi (green) and Corduba (blue)}
\label{frequency}
\end{figure}

@@

We analysed a dataset of 2052 stamps placed in \textit{Germania}. All the data was also compiled by CEIPAC database. We finally selected a total of 1621 stamps consisted of 850 different stamps from 46 sites (see Fig. \ref{germania}). As previously before, the centres were selected with more or equal than five stamps.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/germania5_2.pdf}
\caption{Distribution of the sites in \textit{Germania}. Most sites were located in limes during the Roman Empire}
\label{germania}
\end{figure}

\section{Method: Measuring the dissimilarity}

%\xavi{en stamps o en yacimientos? o en provincias?}

@@

%\xavi{pero este es el único método que usas?}

@@

The analysis shows that amphoric stamps could be correlated with spatial distance. The correlation coefficients range from a minimum to a maximum. The dendrogram in Fig. \ref{dendro} was obtained with Morisita-Horn index. The dendrogram suggests that amphora workshops used different stamps for their production system. Nearby workshops show similarity on the stamps while most of them seem to display different stamps roles. Groups of workshops sharing similar stamps were not found in the cluster: the majority of stamp grouping was composed of no more than three workshops. Additionally, workshops that shared more similar amphoric stamps belonged to the same \textit{conventus} area, such as Picachos, Cerro de los Pesebres and El Castillejo.

\begin{sidewaysfigure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendro}
\caption{Dendrogram obtained by Morisita-Horn algorithm of different amphora workshops in \textit{Baetica} area. Colours are represented by areas divided into \textit{Hispalis} (red), \textit{Astigi} (green) and \textit{Corduba} (blue)}
\label{dendro}
\end{sidewaysfigure}

\subsection{Consumption centres: Britannia and Germania}
%\xavi{mejor consumption?}

@@

\begin{sidewaysfigure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendrobrit5.pdf}
\caption{Dendrogram obtained by Morisita-Horn algorithm of different sites in \textit{Britannia}. Colours are represented by type of sites divided into military sites (red), civil sites (green) and no specific (grey)}
\label{britmap}
\end{sidewaysfigure}

A minor correlation can be explained by the fact that \textit{Britannia} spatial area is much wide than \textit{Baetica} where the centres were more concentrated.

Most military sites that showed greater similarity in the results were geographically close. We identified two factors in the results: 1) nearby sites tend to share similar stamps and 2) most sites have different stamps. 
By contrast, most sites did not show a strong similarity in stamp correlation. Neither we found a grouping of similar stamps in a specific place. In general, we did not observe any production pattern that indicates a clear organisation between production centres and consumption centres. In other words, our results did not indicate the presence of a specific production center from \textit{Baetica} in \textit{Britannia} province. Rather, it seems that olive oil production was distributed by non-specific production centres.

The similarity of stamps shows results less significant than \textit{Britannia} province as it can be seen in the dendrogram (Fig. \ref{germap}).

\begin{sidewaysfigure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendroger5.pdf}
\caption{Dendrogram obtained by Morisita-Horn algorithm of different sites in \textit{Germania}. Colours are represented by types of sites divided into military sites (red), civil sites (green) and no specific (grey)}
\label{germap}
\end{sidewaysfigure}

The sites with more similarity in stamps were closely located in limes. In this case, most sites have different stamps which means that we did not detect any pattern. On the other hand, there seems to be a greater concentration of the seals in areas eminently militarised and close to the German limes. However, It is still no possibility to determine the existence of a defined pattern that reflects in more detail the route of this production in the area of the German limes.

@@

The analysis suggests that there is  a connection between the same stamp code and nearby amphora workshops, excluding certain exceptions. Consequently, the majority of stamps are located in different amphora workshops and only similar stamps between closer amphora workshops were found. Indeed, our results show that most similar stamps were detected in the same \textit{conventus} area. This results can indicate a grouping of stamps with similar codes in the same \textit{conventus}. However, these stamps tend to share the same area of production but there is not a general relation between groups of amphora workshops and connection areas. In other words, we do not identify groups of more than three workshops or less sharing the same amphoric stamps as the dendrogram showed. In general, the majority of stamps were located in different amphora workshops.

@@

This could be defined by several factors. First, each workshop had a different organisation involved in the use of stamps and they were not used in other workshops. Second, stamp similarity in closer workshops could be linked to a spatial pattern. It is more probably than closer workshops tend to share more traits than distant workshops. While the role of the river was significant for the distribution of amphorae in consumption places, river connection amongst workshops does not seem to show relevancy for the distribution of stamps. Finally, the distribution of stamps could have shown some research bias. In some cases, workshops have been catalogued with different names in spite of belonging to the same workshops or being closer between each other. Additionally, most of the workshops were not widely excavated.

@@

It is worth to mention that centres with a similarity correlation were located close to the Atlantic coast around the North Sea and the Celtic Sea. This could indicate that the Atlantic route could have played an essential role in transporting olive oil to the area of \textit{Britannia}, since in the places where there is greater similarity they are found in different strategic points near the sea.

This fact seems to concur with the latest research that suggest the idea of a more important Atlantic route that would connect the provinces militarised\citep{remesal_annona_1986,
remesal_provincial_2008,
carreras_atlantic_2012,
morillo_hispania_2016,rubio-campillo_provincias_2018}.

@@

It is worth to mention that it was not detected a robust model of organisation in both cases with regard to the distribution of stamps in consumption centres. This means
that it is unknown whether certain production centres went to one province or another or, at the
less, that it can be clearly reflected in the data with a greater similarity in the amphoric stamps. Thus, production centres could have distributed randomly olive oil both \textit{Britannia} and \textit{Germania}. We do not detect production centres dedicated to the distribution of olive oil in a certain province.

In general, we observed a distribution of the stamps to military stamps. By contrast, judging by the results obtained, there does not seem to be a specific pattern in terms of geographical distribution. Nor it was detected consumption areas where stamps are specified from an amphora workshop.

\section{Conclusion}

The results of the case study could be interpreted by several reasons. On the one hand, the use of these amphoric stamps could have been exclusively running by the owner or owners of the workshop to distinguish the amphora workshop. This hypothesis would explain the fact that we do not find similar stamps in different workshops; however, we found different amphora stamps in the same workshops that they would be barely difficult to assign different owners.

On the other hand, the existence of different stamps in the same workshop but which are only repeated in very close places it would imply some kind of organisation from within workshop affecting the closest centres. This could be explained such as batch systematic organisation between potters \citep{juanmorostesis}. 
This method allowed to potters organise the production with the batch stamping for the posterior commercialisation.

>>>>

The aim of this study is to explore economics dynamics in the production and distribution of olive oil trade. 
Our case of study has been focused on the production processes located in \textit{Baetica} province (currently Andalusia) from 1st to 3rd AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. \textit{Baetica} became an important production and distribution centre during the Roman Empire. However, it remains under debate about how this province was organised and whether it could be possible to identify patterns in the olive oil market. Amphoric stamps are used to identify the presence of different groups that might share similar stamps. To achieve this goal, we analyse a set of stamps from two centres: 1) production centres by analysing different workshops in \textit{Baetica} province and 2) two Roman provinces such as \textit{Germania} and \textit{Britannia} as consumption centres. They will be used to detect a connection between the distribution of amphoric stamps and the economic structure in both centres. Here, we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance.

@@

The role of the Mediterranean sea for the trade had an enormous impact during the Roman Empire. The spread of thousands of different goods along maritime shipping allowed progressive exploitation of communication networks and a major interaction between communities \citep{rodriguez_baetican_1998, 
temin_market_2001,
bevan_mediterranean_2014}. The exchange of different products between regions can be seen in both archaeological and written records as a frequent indicator of Roman trade.

@@

Given this fact, an important mechanism of production control under the Roman government was created spreading by different areas attending to the richness of different places. Roman provinces such as \textit{Baetica} (currently Andalusia, south Spain) became important production centres of olive oil during the Roman Empire. Olive oil was considered the liquid gold since it was used in different aspect of daily life for cooking, hygiene or lighting  \citep{mattingly_d.j._oil_1988}. As a consequence of high demand, this product was distributed in a large number of amphorae across the provinces, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

The understanding of the olive oil production processes has been widely discussed over many decades\citep{rodriguez_economioleicola_1977, Chic_hispania_1997,millet_anforas_1998}. Despite this topic is still subject to debate: what patterns were followed for the distribution of olive oil to the different provinces?
Did each province follow a different pattern for its distribution? Nor the lack of written records has made possible to detect any indication of patterns in the olive oil market.

Advances in the research of the Roman studies have currently led an environment with more diverse commercial dynamics \citep{duncan1982economy,
temin_economy_2006,
quantifyingwilson2009}. The application of different quantitative approaches combined with the archaeological evidence has allowed us partly to deal with the complexity of the Roman production \citep{brughmans_roman_2016,
orengo_seeds_2016,bayesian_2018,
coto-sarmiento_identifying_2018,
rubio-campillo_ecology_2018}.

This paper aims to study the olive oil market connection between provinces by calculating the similarity of stamps. Specifically, our work pretends to detect microeconomic processes focused on a commercial product from a specific province \citep{isaksen_network_2006}. We want to understand the pattern of olive oil production linked to amphora workshops and amphoric stamps used to mark them. We focus here on exploring the economic relation between stamps and amphora production and distribution centres.

Two case studies have been studied in order to analyse the relation between production centres (\textit{Baetica}) and consumption centres (\textit{Britannia} and \textit{Germania}). In the case of \textit{Baetica} province, we want to identify the role of the stamps in the organisation of the workshop; in Roman provinces, our aim is detecting groups of stamps concentrated in an area or if some groups have an important role for the exportation of olive oil in those provinces. This economic connection could be identified by different aspects: a) correlation between spatial distance and centres based on the idea that closer workshops concentrate similar amphoric stamps in a specific area than the distant workshops and b) groups of similar stamps were concentrated in a specific province.

In particular, we study the distribution of amphoric stamps to identify a correlation between geographical distance and similarity. Based on this assumption, we proposed three hypotheses: a) we can identify a correlation between spatial distance and the distribution of stamps, b) stamps located in close workshops share similar traits and c) low mobility of amphoric stamps to other regions: stamps always stay in the same region.

@@

The high demand of Roman provinces stimulated by the good condition of the \textit{Baetica} lands allowed to develop a massive infrastructure of olive oil production. The production and distribution of olive oil were growing exponentially for almost three centuries \citep{remesal_concierto}. As a result, hundreds of amphora workshops were raised to support this high demand of the Roman Empire. The amphora workshops were situated in different locations in \textit{Baetica} province, along the rivers Guadalquivir and Genil (see Fig.\ref{workshop}).

The strategic location of the workshops close to the river allowed the transport of olive oil through river connectivity to the routes of the Mediterranean and the Atlantic \citep{garcia_vargas_enrique_formal_2010}.

@@

\section{A potential indicator of Roman economy: Dressel 20 stamps}

@@

Most stamps display a large activity of production that it can be difficult for specifying an accurate chronology in some cases. They have been frequently dated with the consular dating by studying \textit{tituli picti} found in \textit{Monte Testaccio} \citep{Testaccio1, berni_millet_epigrafianforica_2008}.
However, the chronology of stamps could be also biased in other sites taking into account that amphorae were deposited after being produced and consumed.

@@

Frequently, they were mainly marked in handles, but seldom in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion. Stamps are mostly formed by a code of three letters and they can appear in abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.

Despite researchers identify stamps as identity marks, there is not a consensus about the meaning of the stamps \citep{rodriguez_baetican_1998}. Alike a high percentage of Dressel 20 containers were not stamped being difficult to determinate.

@@

Nevertheless, some challenges remain under discussion such as how this production was organised and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow a distribution pattern? Did stamps share the same workshop?

Neither the use of written records have allowed providing enough information that can explain the economic role of \textit{Baetica} province in the Roman organisation.

@@

The conquest of new provinces allowed the Roman Empire the arrival of resources through Mediterranean and Atlantic routes. This lead to a progressive change in the economic and social structure creating new trade networks for supplying Roman conquest. As a result, Augusto's administration created the figure of the \textit{praefectura annonae} for the supply of wheat. The role of the \textit{praefectura annonae} was mainly to provide though \textit{frumentationes} a quantity per month of wheat to Roman citizens \citep{remesal_annona_1986,remesal_concierto}.

@@

The military consumption of olive oil in Roman provinces may be related to two aspects: a) cultural consumption whose product is consumed by cultural reasons such as to identity or habit and b) economical reasons where olive oil distribution is based on the transport costs, unlike other products \citep[69-70]{carreras_britannia_1998}.

@@

At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}. This fact will have a spatial incidence in sites close at Hadrian Wall.  
Olive oil production in \textit{Baetica} would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic points \citep{carreras_atlantic_2012}. The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. Thus, the network was mainly focused on the support of soldiers during military campaigns.

@@

Finally, the increase in olive oil exports would experience a progressive slowdown from the third century A.D., coinciding with the change in market strategy in the Empire. As a result, a progressive decrease of Dressel 20 is documented being gradually replaced for Dressel 23 amphorae \citep{rodriguez1991aceite,millet_anforas_1998}.

@@

The Roman studies about the presence of Dressel 20 amphorae in \textit{Germania} has not had the same repercussion as the rest of the province. This could be explained due to the lack of archaeological sources \citep{horacio2010llegada}.

@@

As the case of \textit{Britannia}, the consumption of olive oil in \textit{Germania} was not detected before the conquest. This is evidenced by the presence of Dressel 20 amphora in different sites close to \textit{limes}.

The presence of the Roman army encouraged the exchange in the province showed by the arrival of this product both civil settlement and military sites with a mayor concentration at German \textit{limes}. It seems that some Baetican centres would be assigned to the support of olive oil. However, this hypothesis can not be proven by the lack of archaeological sources \citep[125]{remesal_concierto}.

\section{Material and Methods}

The goal of this study is exploring the effect of the production patterns between different centres. We are especially interested in identifying links between production and consumption centres by using amphoric stamps. To do that, we use the CEIPAC database to collect stamps from different places. The CEIPAC dataset contains over 50.000 of epigraphy records found in amphorae, mostly from \textit{Monte Testaccio}.

Therefore, this study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. A way to analyse is to use a quantitative framework to measure the similarity between amphoric stamps. Here we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potential spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance.

@@

We studied a dataset of 3798 stamps collected from different Dressel 20 amphora workshops in \textit{Baetica} province. The database was compiled by CEIPAC group\citep{remesal_centro_2015} (see CEIPAC database here \url{http://romanopendata.eu}). However, approximately 70 \% of stamps cannot be tested due to the fragmentation of the dataset. Consequently, any stamps with incomplete information were discarded and not integrated into our dataset. We finally filtered a total sample of 987 stamps composed of 130 different stamps from 81 workshops.

@@

From the database, we collected the site where stamps were found and the stamp code. We also created a new column with the area of the stamps. These areas, known as \textit{conventus}, were administrative centres for territorial organisation in the Roman Empire \citep[58]{ozcariz_gil_administracion_2013}.
Dressel 20 stamps were found in three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter Hispalis), \textit{Cordubensis} (currently C\'ordoba, hereafter Corduba) and \textit{Astigi} (currently Écija, Sevilla, hereafter Astigi) \citep{rodriguez_economioleicola_1977,
chicdatos2001,berni_millet_epigrafianforica_2008} . It is important to mention that some workshops exhibited the same coordinates but they were differently catalogued as a different workshop. 
\maria{seguir con esta mierda}
We calculated the frequency of distribution in amphora stamps with the purpose of implementing an exploratory analysis (EDA) to detect the distribution and the number of stamps per each centre.

The frequency of stamps per workshop can be seen in Fig. \ref{stamps}. Most 
workshops contained only one stamp whereas one workshop (La Catria) concentrated a high frequency of amphoric stamps with a total number of 228 stamps. The type of distribution is also frequent in amphora production where we observed a self-organised complex system pattern with a major concentration of the number of stamps in a few workshops \citep{bayesian_2018,coto-sarmiento_identifying_2018}.

@@

The distribution of amphora stamps in different \textit{conventus} can be seen in Fig. \ref{frequency}. The majority of stamps are concentrated in \textit{Hispalis} with 574 stamps while \textit{Corduba} and \textit{Astigi} with 267 and 146 stamps, respectively. Mostly, workshops show a homogeneity on the frequency of stamps except for both La Catria and Arva that exhibit a big amount of stamps with 29 different code stamps. According to previous studies, those workshops became in most important centres of amphora production although they could have more intensely prospected than others. \citep{arva_1997}.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/frequency}
\caption{Distribution of the number of different code stamps (X axis) for each \textit{conventus} (Y axis). Each dot corresponds to a workshop sorted by different areas. Colours are represented by areas divided into Hispalis (red), Astigi (green) and Corduba (blue)}
\label{frequency}
\end{figure}

@@

We analysed a dataset of 2052 stamps placed in \textit{Germania}. All the data was also compiled by CEIPAC database. We finally collected a total of 1621 stamps consisted of 850 different stamps from 46 sites (see Fig. \ref{germania}). As previously before, the centres were selected with more or equal than five stamps.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/germania5_2.pdf}
\caption{Distribution of the sites in \textit{Germania}. Most sites were located in German \textit{limes}}
\label{germania}
\end{figure}

\subsection{Measuring the dissimilarity}

@@

@@

%The correlation coefficients range from a minimum to a maximum.

The analysis shows that the similarity of amphoric stamps could be correlated with spatial distance. The dendrogram in Fig. \ref{dendro} was obtained with Morisita-Horn index. The dendrogram suggests that amphora workshops used different stamps for their production system. Nearby workshops show similarity on the stamps while most workshops seem to display different stamps. Groups of workshops sharing similar stamps were not found in the dendrogram: the majority of stamp grouping was composed of no more than three workshops. Additionally, workshops that shared more similar amphoric stamps belonged to the same \textit{conventus} area, such as Picachos, Cerro de los Pesebres and El Castillejo.

\begin{sidewaysfigure}[htp]
	\centering
\includegraphics[angle=180, width=\linewidth]{figs/dendro}
\caption{Dendrogram obtained by Morisita-Horn algorithm of different amphora workshops in \textit{Baetica} area. Colours are represented by areas divided into \textit{Hispalis} (red), \textit{Astigi} (green) and \textit{Corduba} (blue)}
\label{dendro}
\end{sidewaysfigure}

@@

\begin{sidewaysfigure}[htp]
	\centering
\includegraphics[angle=180,width=\linewidth]{figs/dendrobrit5.pdf}
\caption{Dendrogram obtained by Morisita-Horn algorithm of different sites in \textit{Britannia}. Colours are represented by type of sites divided into military sites (red), civil sites (green) and no specific (grey)}
\label{britmap}
\end{sidewaysfigure}

A minor correlation can be explained by the fact that \textit{Britannia} spatial area is much wider than \textit{Baetica} where the centres were more concentrated.

Most military sites that showed a high similarity in the results were geographically close. We identified two factors: 1) nearby sites tend to share similar stamps and 2) most sites have different stamps. 
By contrast, most sites did not show a strong similarity in stamp correlation. Neither we found a grouping of similar stamps in a specific place. In general, the dendrogram did not observe any production pattern that indicates a clear organisation between production centres and consumption centres. In other words, our results did not indicate the presence of a specific production centre from \textit{Baetica} in \textit{Britannia} province. Rather, it would imply that olive oil production was distributed by non-specific production centres.

In \textit{Germania}, the similarity of stamps shows results less significant than \textit{Britannia} province as it can be seen in the dendrogram (Fig. \ref{germap}).

\begin{sidewaysfigure}[htp]
	\centering
\includegraphics[angle=180, width=\linewidth]{figs/dendroger5.pdf}
\caption{Dendrogram obtained by Morisita-Horn algorithm of different sites in \textit{Germania}. Colours are represented by types of sites divided into military sites (red), civil sites (green) and no specific (grey)}
\label{germap}
\end{sidewaysfigure}

A higher concentration of sites sharing similar stamps were found in areas eminently militarised and close to German \textit{limes}, even if most sites mostly showed different stamps.
The interpretation of the results suggests that there is still no possibility to determine the existence of a defined pattern that reflects in more detail the route of this production in the area of the German limes.

@@

The analysis suggests a connection between the same stamp code and nearby amphora workshops, excluding certain exceptions. Consequently, the majority of stamps are located in different amphora workshops and only similar stamps between closer amphora workshops were found. Indeed, our results show that most similar stamps were detected in the same \textit{conventus} area. This results may indicate a grouping of stamps with similar codes in the same \textit{conventus}. Despite these stamps tend to share the same area of production, we do not identify groups of any more than three workshops sharing the same amphoric stamps as the dendrogram showed. In general, the majority of stamps were located in different amphora workshops.

@@

This could be defined by several factors. First, each workshop had a different organisation involved in the use of stamps and they were not used in other workshops. Second, stamp similarity in closer workshops could be linked to a spatial pattern. It is more probably than closer workshops tend to share more traits than distant workshops. While the role of the river was significant for the distribution of amphorae, river connection amongst workshops does not seem to show relevancy for the distribution of stamps. Finally, the distribution of stamps could have shown some research bias. In some cases, workshops have been catalogued with different names in spite of belonging to the same workshops or being closer between each other. Additionally, most of the workshops were not widely excavated.

@@

It should be noted that centres with a similarity correlation were located close to the Atlantic coast around the North Sea and the Celtic Sea. This may indicate that the Atlantic route could have played an essential role in transporting olive oil to the area of \textit{Britannia}, since in the places where there is greater similarity they are found in different strategic points near the sea.

This fact seems to concur with the latest research that suggests the idea of a more important Atlantic route that would connect the militarised provinces \citep{remesal_annona_1986,
remesal_provincial_2008,
carreras_atlantic_2012,
morillo_hispania_2016,rubio-campillo_provincias_2018}.

@@

It is worth to mention that it was not detected a robust model of organisation in both cases with regard to the distribution of stamps in consumption centres. This evidence means that it is unknown whether some production centres went to one province or another or, at least, that they can be clearly reflected in the data with a greater similarity in the amphoric stamps. Thus, production centres could have distributed randomly olive oil both \textit{Britannia} and \textit{Germania}. Neither we do not detect production centres dedicated to the distribution of olive oil in a certain province.

%In general, we observed a concentration of the stamps in military centres.

Judging by the results obtained, there does not seem to be a specific pattern in terms of geographical distribution. Nor it was detected consumption areas where stamps are specified from an amphora workshop.

\section{Concluding remarks}

The results of this work can be interpreted for several reasons. On the one hand, the use of amphoric stamps could have been exclusively running by the owner or owners of the workshop to distinguish the amphora workshop. This hypothesis would explain the fact that we do not find similar stamps in different workshops; however, different code stamps have been detected in the same workshops that they would be barely difficult to assign to different owners.

On the other hand, the presence of different stamps in the same workshop could imply some kind of organisation from within the workshop affecting the closest centres \citep{juanmorostesis}. The marked amphorae could have allowed to potters organise the production with a batch stamping for the posterior commercialisation.

This could be explained such as batch systematic organisation between potters \citep{juanmorostesis}. 
This method allowed to potters organises the production with the batch stamping for the posterior commercialisation.
